---
title: "Grammar of Data Manipulation in R"
author: "A. Jordan Nafa"
date: "February 9, 2023"
institute: "University of North Texas"
subtitle: "PSCI 3300.002 Political Science Research Methods"
format:
  revealjs:
    theme: blood
    highlight-style: monokai
    toc-depth: 2
    embed-resources: true
    citations-hover: true
    code-link: true
    transition: "fade"
    code-block-bg: "#272822"
    logo: "../../../../assets/logo/psci-3300-sp23-dag-trimmed-hex.png"
    css: "../../../../assets/css/slide-style.css"
    include-in-header: "../../../../assets/math-colors.js"
editor: visual
width: 1360
height: 800
bibliography: "../../../../assets/references.bib"
csl: "../../../../assets/apsa.csl"
link-citations: yes
---

```{r, include=FALSE}
## Load required packages
pacman::p_load(
  "tidyverse",
  "dagitty",
  "ggraph",
  "ggdag",
  install = FALSE
  )

## Base theme for the figures
fig_theme <- theme_light(base_size = 24, base_family = "serif") +
  ## Settings sepcific to the reveal.js theme
  theme(
    strip.background = element_blank(),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA),
    plot.title = element_text(color = "white"),
    plot.subtitle = element_text(color = "white"),
    plot.caption = element_text(color = "white", face = "italic"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white", face = "bold"),
    legend.title = element_text(color = "white", face = "bold"),
    legend.text = element_text(color = "white", face = "bold"),
    strip.text = element_text(face = "bold"),
    plot.caption.position = "plot",
    legend.position = "top"
  )

shorten_dag_arrows <- function(tidy_dag, proportion){
  # Update underlying ggdag object
  tidy_dag$data <- dplyr::mutate(tidy_dag$data, 
                                 xend = (1-proportion/2)*(xend - x) + x, 
                                 yend = (1-proportion/2)*(yend - y) + y,
                                 xstart = (1-proportion/2)*(x - xend) + xend,
                                 ystart = (1-proportion/2)*(y-yend) + yend)
  return(tidy_dag)
}
```

# Principles of Tidy Data

## A Grammar of Data Manipulation

::: incremental
-   Data grammar and `{tidyverse}`

    -   **verbs** are functions that that do things

        -   `select`, `filter`, `mutate`, `transmute`, `summarize`, `arrange`, `left_join`, `inner_join`, and many more in `{dplyr}` and `{tidyr}`

    -   **nouns** are data frames, the things verbs act upon

    -   The **tidyverse** is a collection of data manipulation and visualization packages

        -   `{ggplot2}`, `{purrr}`, `{tibble}`, `{dplyr}`, `{tidyr}`, `{stringr}`, `{readr}`, and `{forcats}` are part of the `{tidyverse}` package

-   For more on this read the chapter on data transformation in [R for Data Science](https://r4ds.had.co.nz/transform.html) or see the [dplyr cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-transformation.pdf)
:::

## Reading Data into R

::: incremental
-   To read external data into R, we can use the `read_csv` for standard tab delimited data formats or `read_rds` for the native R rds format from the `{readr}` package

-   For proprietary file formats like those used by Stata, SPSS, and SAS we can use functions such as `read_dta`, `read_spss`, and `read_sas` from the `{haven}` package

-   The `{arrow}` package provides an R interface to Apache Arrow that supports common dplyr verbs

    -   Allows for larger-than-memory data analytics

    -   Provides access to the parquet file storage format via the `read_parquet` and `write_parquet` formats
:::

## Reading Data into R

-   The first step is to load the `{tidyverse}` package which we can do using either `library` or `pacman::p_load`

## Reading Data into R

-   The first step is to load the `{tidyverse}` package which we can do using either `library` or `pacman::p_load`

```{r, echo=TRUE}
# Load {tidyverse} using library()
library(tidyverse) # Suite of data management packages
```

## Reading Data into R

-   The first step is to load the `{tidyverse}` package which we can do using either `library` or `pacman::p_load`

```{r, echo=TRUE}
# Load {tidyverse} using library()
library(tidyverse) # Suite of data management packages
```

-   Next we'll read in a subset of the Integrated Values Survey [@WVSTrend; @EVSTrend] which you can either download from Canvas or retrieve using `git pull` in the RStudio terminal

## Reading Data into R

-   The first step is to load the `{tidyverse}` package which we can do using either `library` or `pacman::p_load`

```{r, echo=TRUE}
# Load {tidyverse} using library()
library(tidyverse) # Suite of data management packages
```

-   Next we'll read in a subset of the Integrated Values Survey [@WVSTrend; @EVSTrend] which you can either download from Canvas or retrieve using `git pull` in the RStudio terminal

```{r, eval=FALSE, echo=TRUE}
# Read in the IVS Data
ivs <- read_rds("data/IVS_Data_Subset.rds")
```

```{r, echo=FALSE}
# Read in the IVS Data
ivs <- read_rds(str_c(here::here(), "/data/IVS_Data_Subset.rds"))
```

## Reading Data into R

-   We can use `glimpse` to print a preview of the data in a readable format

## Reading Data into R

-   We can use `glimpse` to print a preview of the data in a readable format

```{r, echo=TRUE}
# Print a preview of the data
glimpse(ivs)
```

## Reading Data into R

-   We can use `glimpse` to print a preview of the data in a readable format

```{r, echo=TRUE}
# Print a preview of the data
glimpse(ivs)
```

-   We see the data has 40 columns and 364,771 rows, each of which corresponds to a unique survey respondent

## Transforming Data

::: incremental
-   We'll cover the major `{dplyr}` verbs today

    -   `mutate` transforms a column or columns in a data frame in place

        -   It either adds a new column or replaces an existing one and returns the modified data frame

    -   `select` performs column-wise subsetting on a data frame

    -   `transmute` transforms a column or columns in a data frame and returns only the transformed columns

        -   `transmute` is like `select` and `mutate` combined into a single function

    -   `filter` performs row-wise subsetting on a data frame

    -   `arrange` sorts the data by values in the specified column(s)
:::

## The Pipe

::: incremental
-   The logic of the pipe operator is simple in that the default behavior is to pass some object, in this case a data frame, to the first argument of a function $f(x,y)$ which we can express as

    -   $\mathrm{data} \enspace \%\gt\% \enspace function(., \, y) = function(\mathrm{data}, \, y)$

-   More complex formulations are also possible thanks to the . operator which allows us to pass an object to higher order arguments in a function as in the case of the expression

    -   $\mathrm{data} \enspace \%\gt\% \enspace function(x, \, y, \,.) = function(x, \, y, \,\mathrm{data})$
:::

## Transforming Data {.auto-animate}

-   `mutate` transforms a column or columns in a data frame in place and returns the original data frame

## Transforming Data {.auto-animate}

-   `mutate` transforms a column or columns in a data frame in place and returns the original data frame

-   Here we are creating a new column `female` in the data frame `ivs` and using `if_else` to assign it a 1 if a respondent is female based on the existing column `sex` and 0 otherwise

## Transforming Data {.auto-animate}

-   `mutate` transforms a column or columns in a data frame in place and returns the original data frame

-   Here we are creating a new column `female` in the data frame `ivs` and using `if_else` to assign it a 1 if a respondent is female based on the existing column `sex` and 0 otherwise

``` r
ivs <- ivs %>% 
  mutate(
    # Add a column to the ivs data frame object that recodes respondent sex
    female = if_else(sex == "Female", 1,  0),
  )
```

## Transforming Data {.auto-animate}

-   `mutate` transforms a column or columns in a data frame in place and returns the original data frame

-   Here we are creating a new column `female` in the data frame `ivs` and using `if_else` to assign it a 1 if a respondent is female based on the existing column `sex` and 0 otherwise

``` r
ivs <- ivs %>% 
  mutate(
    # Add a column to the ivs data frame object that recodes respondent sex
    female = if_else(sex == "Female", 1,  0),
    # Add a column to the ivs data frame that takes a value of 1 for surveys in 2000, 2010, 2020
    decade_start = if_else(year %in% c(2000, 2010, 2020), 1, 0)
  )
```

## Transforming Data {.auto-animate}

-   `mutate` transforms a column or columns in a data frame in place and returns the original data frame

-   Here we are creating a new column `female` in the data frame `ivs` and using `if_else` to assign it a 1 if a respondent is female based on the existing column `sex` and 0 otherwise

``` r
ivs <- ivs %>% 
  mutate(
    # Add a column to the ivs data frame object that recodes respondent sex
    female = if_else(sex == "Female", 1,  0),
    # Add a column to the ivs data frame that takes a value of 1 for surveys in 2000, 2010, 2020
    decade_start = if_else(year %in% c(2000, 2010, 2020), 1, 0)
  )
```

-   We can use the `%in%` operator to match multiple conditions in the same column. `year %in% c(2000, 2010, 2020)` is equivalent to `year == 2000 | year == 2010 | year == 2020`

## Subsetting Data {.auto-animate}

-   `select` performs **column-wise subsetting** on a data frame

## Subsetting Data {.auto-animate}

-   `select` performs **column-wise subsetting** on a data frame

``` r
ivs_sub <- ivs %>% 
  # select extracts only the columns we specify
  select(country_jj:project, soc_pctfemleg, pctfemleg, churchill)
```

## Subsetting Data {.auto-animate}

-   `select` performs **column-wise subsetting** on a data frame

``` r
ivs_sub <- ivs %>% 
  # select extracts only the columns we specify
  select(country_jj:project, soc_pctfemleg, pctfemleg, churchill)
```

-   Here we extracting the columns in the range `country_jj` through `project` using the `:` operator along with `soc_pctfemleg`, `pctfemleg`, and `churchill`

## Subsetting Data {.auto-animate}

-   `select` performs **column-wise subsetting** on a data frame

```{r, echo=TRUE}
ivs_sub <- ivs %>% 
  # select extracts only the columns we specify
  select(country_jj:project, soc_pctfemleg, pctfemleg, churchill)
```

-   Here we extracting the columns in the range `country_jj` through `project` using the `:` operator along with `soc_pctfemleg`, `pctfemleg`, and `churchill`

```{r, echo=TRUE}
# Print the new object
glimpse(ivs_sub)
```

## Filtering Data {.auto-animate}

-   `filter` performs **row-wise subsetting** on a data frame and returns all **rows** that match the specified condition

## Filtering Data {.auto-animate}

-   `filter` performs **row-wise subsetting** on a data frame and returns all **rows** that match the specified condition

``` r
ivs_sub_decades <- ivs_sub %>% 
  # filter subsets the rows that match a specified condition
  filter(year %in% c(2000, 2010, 2020))
```

## Filtering Data {.auto-animate}

-   `filter` performs **row-wise subsetting** on a data frame and returns all **rows** that match the specified condition

``` r
ivs_sub_decades <- ivs_sub %>% 
  # filter subsets the rows that match a specified condition
  filter(year %in% c(2000, 2010, 2020))
```

-   Here we are extracting the rows that contain 2000, 2010, or 2020 in the column `year`

## Filtering Data {.auto-animate}

-   `filter` performs **row-wise subsetting** on a data frame and returns all **rows** that match the specified condition

```{r, echo=TRUE}
ivs_sub_decades <- ivs_sub %>% 
  # filter subsets the rows that match a specified condition
  filter(year %in% c(2000, 2010, 2020))
```

-   Here we are extracting the rows that contain 2000, 2010, or 2020 in the column `year`

```{r, echo=TRUE}
# Print the new object
glimpse(ivs_sub_decades)
```

## Sorting Data {.auto-animate}

-   `arrange` sorts the data by values in the specified column(s)

## Sorting Data {.auto-animate}

-   `arrange` sorts the data by values in the specified column(s)

```{r, echo=TRUE}
ivs_sub_decades <- ivs_sub_decades %>% 
  # sort the data in ascending order by country and year 
  arrange(country_jj, year)

# Print the new object
glimpse(ivs_sub_decades)
```

## Summarizing Data {.auto-animate}

-   `summarize` can be used in conjunction with `group_by` to perform aggregation and obtain grouped summary statistics and cross-tabulations

## Summarizing Data {.auto-animate}

-   `summarize` can be used in conjunction with `group_by` to perform aggregation and obtain grouped summary statistics and cross-tabulations

``` r
ivs_sub_summ <- ivs_sub_decades %>% 
  # Group the data by country and survey
  group_by(country_jj, year, project) %>% 
  # Summarize the data, taking the mean and sd
  summarise(across(
    .cols = c(soc_pctfemleg, churchill),
    .fns = list(
      mean = ~ mean(.x, na.rm = TRUE),
      sd = ~ sd(.x, na.rm = TRUE)
    )
  ))
```

## Summarizing Data {.auto-animate}

-   `summarize` can be used in conjunction with `group_by` to perform aggregation and obtain grouped summary statistics and cross-tabulations

```{r, echo=TRUE}
ivs_sub_summ <- ivs_sub_decades %>% 
  # Group the data by country and survey
  group_by(country_jj, year, project) %>% 
  # Summarize the data, taking the mean and sd
  summarise(across(
    .cols = c(soc_pctfemleg, churchill),
    .fns = list(
      mean = ~ mean(.x, na.rm = TRUE),
      sd = ~ sd(.x, na.rm = TRUE)
    )
  ))
```

-   For each column in the range `soc_pctfemleg:churchill`, we take the mean and the standard deviation within survey-years and assign it to an object `ivs_sub_summ`

## Licensing and Copyright Information

-   All text and images in this course are made available for public non-commercial use under a [Creative Commons Attribution-ShareAlike 4.0 International (CC BY-SA 4.0) License](https://creativecommons.org/licenses/by-sa/4.0/).

-   All R, HTML, and CSS code is provided for public use under a [BSD 3-Clause License](https://github.com/ajnafa/PSCI-3300-Political-Research-Methods-SP23/blob/main/LICENSE).

-   The files and code necessary to reproduce the content of this course are or will be made available via [the course's github repository](https://github.com/ajnafa/PSCI-3300-Political-Research-Methods-SP23) with the exception of those covered by existing commercial copyright restrictions (i.e., copies of the assigned readings for the course).

## References
